---
title: "EDS241: Assignment 1"
author: "Briana Barajas"
date: '`r format(Sys.time(), "%m/%d/%Y")`'
output: 
  pdf_document:
    toc: false
    number_sections: yes
header-includes:
  - \setlength{\parindent}{1em}
  - \usepackage{float}
---

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}

# set default chunk options
knitr::opts_chunk$set(fig.width = 4, fig.height = 3, 
                      echo = TRUE, message = FALSE, warning = FALSE)

# load packages
packages=c("MASS", "tidyverse", # Used in assignment 1
           "stargazer", "here", "tidyr", "dplyr","stringr", "janitor", # Used for Mock assignment
           "cowplot", "ggplot2", "tinytex", "datasets", "tibble") # Used for Mock assignment

for (i in packages) {
  if (require(i,character.only=TRUE)==FALSE) {
    install.packages(i,repos='http://cran.us.r-project.org')
  }
  else {
    require(i,character.only=TRUE)
  }
}

# Disable scientific notation if you want
options(scipen=999)

```

# Part 1

(NOTE: Uses the RCT.R code provided with lecture to generate data) DO NOT CHANGE ANYTHING BELOW UNTIL IT SAYS EXPLICITLY)

```{r , include=FALSE}

# Generate data on potential outcomes and pre-treatment covariates:

rm(list=ls())
library(MASS)

# Define variables and generate data:
N <- 20000
Xi <- sample(c(1,2,3,4,5),N, replace=TRUE)
m0_Xi <- 0.5*Xi
m1_Xi <- 1*Xi

# Generate correlated error terms:
ei <- mvrnorm(n=N,mu=c(0, 0),Sigma=matrix(c(1,0.75,0.75,1), ncol=2))
# Calculate potential outcomes:
Yi_0 = m0_Xi + ei[,1]		
Yi_1 = m1_Xi + ei[,2]

# Output the mean of the potential outcomes:
mean(Yi_0)
mean(Yi_1)

# Create a dataframe from the vectors:
df <- data.frame(Xi, Yi_0, Yi_1)

```

## BELOW YOU CAN (AND HAVE TO) CHANGE AND ADD CODE TO DO ASSIGNMENT

**Part 1**: Use the small program above that generates synthetic potential outcomes without treatment, Yi_0, and with treatment, Yi_1. When reporting findings, report them using statistical terminology (i.e. more than y/n.) Please do the following and answer the respective questions (briefly).

a)  Create equally sized treatment and control groups by creating a binary random variable Di where the units with the "1's" are chosen randomly.

    ```{r}
    # set seed to keep random column the same every time code runs
    set.seed(123)

    # calculate 1/2 the population
    N_half <- N/2

    # create Di column w/proportionate values of 0 and 1, randomly assigned
    df$Di <- sample(c(rep(1, N_half), rep(0, N_half)))
    head(df, 2)
    ```

b)  Make two separate histograms of Xi for the treatment and control group. What do you see and does it comply with your expectations, explain why or why not?

    ```{r}
    ggplot(data = df, mapping = aes(x = Xi)) +
      geom_histogram(aes(fill = Di)) +
      labs(y = 'Count') +
      facet_wrap(~Di, ncol = 1,
                 labeller = labeller(Di = 
        c("0" = "Control (Di = 0)",
          "1" = "Treatment (Di = 1)"))) +
      scale_fill_binned(high = 'maroon', low = 'dodgerblue') +
      guides(fill = 'none') 

    ```

    **ANS:** The histogram displays the expected results, there is a relatively even distribution between the treatment and control groups. Both groups also have an even number of units.

c)  Test whether Di is uncorrelated with the pre-treatment characteristic Xi and report your finding.

    ```{r}
    # print correlation results 
    cor.test(df$Di, Xi)
    ```

    **ANS:** The correlation value between $D_i$ and $X_i$ is a very small, and the p-value is large, so we fail to reject the null hypothesis that the true correlation is equal to 0. In other words, $D_i$ is uncorrelated with the pre-treatment characteristic $X_i$.

d)  Test whether Di is uncorrelated with the potential outcomes Yi_0 and Yi_1 and report your finding (only possible for this synthetic dataset where we know all potential outcomes).

    ```{r}
    cor.test(df$Di, df$Yi_0)
    cor.test(df$Di, df$Yi_1)
    ```

    **ANS:** The correlation between $D_i$ and $Y_{i0}$ is slightly stronger compared to $Y_{i1}$, but overall there is no correlation between treatment ($D_i$) and $Y_{i0}$ or $Y_{i1}$.

e)  Estimate the ATE by comparing mean outcomes for treatment and control group. Test for mean difference between the groups and report your findings.

    ```{r}
    df %>% summarise(ATE = mean(Yi_1 - Yi_0))
    ```

f)  Estimate the ATE using a simple regression of (i) Yi on Di and (ii) Yi on Di and Xi and report your findings.

    ```{r}
    # create a Yi column that will be used for the regressions
    df <- df %>% 
      mutate(Yi = (Di*Yi_1+(1-Di)*Yi_0))

    # regression of Yi on Di
    lm(Yi~Di, data = df)

    # regression of Yi on Di and Xi
    lm(Yi ~ Di + Xi, data = df)
    ```

\newpage

# Part 2

\indent \*\*Part 2\*\* is based on Gertler, Martinez, and Rubio-Codina (2012) (article provided on canvas) and covers impact evaluation of the Mexican conditional cash transfer Progresa (later called Oportunidades, now Prospera). Basically, families with low-incomes received cash benefits if they complied to certain conditions, such as regular school attendance for children and regular healthcare visits. You can read more about the program in the Boxes 2.1 (p.10) & 3.1 (p.40) of the Handbook on impact evaluation: quantitative methods and practices by Khandker, B. Koolwal, and Samad (2010). The program followed a randomized phase-in design. **You have data on households (hh) from 1999, when treatment hh have been receiving benefits for a year and control hh have not yet received any benefits.** You can find a description of the variables at the end of the assignment. Again, briefly report what you find or respond to the questions.

```{r , include=FALSE}
# Load the dataset
progresa <- read_csv("data/hw-1-data.csv")

# change NAs in treatment col to 0
progresa$treatment[is.na(progresa$treatment)] <- 0
```

a)  Some variables in the dataset were collected in 1997 before treatment began. **Use these variables to test whether there are systematic differences between the control and the treatment group before the cash transfer began (i.e. test for systematic differences on all 1997 variables).** Describe your results. Does it matter whether there are systematic differences? Why or why not? Would it be a mistake to do the same test with these variables if they were collected after treatment began and if so why? Note: If your variable is a proportion (e.g. binary variables), you should use a proportions test, otherwise you can use a t-test.

```{r , include=FALSE}

## For continuous variables you can use the t-test
#t.test()

## For binary variables you should use the proportions test
#prop.test()

```

```{r}
# creating subsets for control/treatment groups
control <- progresa %>% 
  filter(treatment == 0)
treated <- progresa %>% 
  filter(treatment == 1)

# continuous variables (t.test)
t.test(hhsize97 ~ treatment, data = progresa)
t.test(vani ~ treatment, data = progresa)
t.test(vani1 ~ treatment, data = progresa)
t.test(vani2 ~ treatment, data = progresa)

# binary variables (proportion test)
a <- table(progresa$treatment, progresa$dirtfloor97)[, c(2,1)]

```

b)  Estimate the impact of program participation on the household's value of animal holdings (vani) using a simple univariate regression. Interpret the intercept and the coefficient. Is this an estimate of a treatment effect?

    ```{r}
    vani_model <- lm(vani ~ treatment, data = progresa)
    summary(vani_model)
    ```

    **ANS:**

c)  Now, include at least 6 independent control variables in your regression. How does the impact of program participation change? Choose one of your other control variables and interpret the coefficient.

    ```{r}
    lm(vani ~ treatment + age_hh + educ_hh + ethnicity_hh + healthcenter + min_dist + hairrigation, data = progresa)
    ```

    **ANS:**

d)  The dataset also contains a variable `intention_to_treat`. This variable identifies eligible households in participating villages. Most of these households ended up in the treatment group receiving the cash transfer, but some did not. Test if the program has an effect on the value of animal holdings of these non-participants (spillover effects). Think of a reason why there might or might not be spillover effects.

    ```{r}
    # find individuals that qualified for AND recieved treatment (1,1)
    table(treatment = progresa$treatment, 
          int_to_treat = progresa$intention_to_treat, exclude = NULL)

    # new treatment variable
    progresa$pseudo_treatment <- ifelse(progresa$intention_to_treat == 1 & progresa$treatment == 0,
                       1, 0)

    # test 
    summary(lm(vani ~ pseudo_treatment, data = progresa))

    ```

\noindent **Hint**: Create a pseudo-treatment variable that is = 1 for individuals who were intended to get treatment but did not receive it, = 0 for the normal control group and excludes the normal treatment group.

```{r , include=FALSE}
# Examine number of hh that were intended to get treatment and that ended up receiving treatment
table(treatment = progresa$treatment, intention_to_treat = progresa$intention_to_treat, exclude = NULL)

# Create a new treatment variable that is:
# = 1 if intention_to_treat == 1 AND not in the actual treatment
# = 0 for normal control hh.

# This creates a new variable called pseudo_treatment that has missing values for every hh
progresa$pseudo_treatment <- NA
# Replace NAs with 1 for the new intervention group
progresa$pseudo_treatment[] <- 1
# Replace NAs with 0 for the normal control group
progresa$pseudo_treatment[] <- 0

```

\newpage
